---
title: "Data Analysis and visualization using R"
output: html_document
---


<a href="/" ><button style="    background: linear-gradient(45deg, #a706b0, #a94442);
    display: inline-block;
    color: #fff;
    border-radius: 5px;
    font-size: 16px;
    padding: 3px 8px;
    outline: none;">Back</button></a>

Overview and purpose
------------------------
<style type="text/css">

body, td {
   font-size: 16px;
   font-family:Georgia;
}
code.r{
  font-size: 12px;
}
pre {
  font-size: 12px
}
</style>




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This article covers the analysis of mpg dataset, create visualization and draw inferences.

### What we would be covering

1. Understanding Data Set
2. Analysing and Building visualisations.


### 1. Understanding Data Set

The data we will be using comes with <strong>" ggplot "</strong>. Below is the breif description of the dataset and its variables.

```{r message=FALSE, warning=FALSE}

library(ggplot2)
library(plotly)

#mpg comes with ggplot2
# change mpg to mpg_sample

mpg_sample<-mpg

#summary of mpg_sample

str(mpg_sample)


```


Out of all the given variable,we are more interested in trans, displ, hwy, cty, cycl, drv.
Each term discription are as follows:
<ol style={{listStyle:"none"}}>
  <li><strong>trans :</strong> Transmission of the vechile which allows it to change gears.</li>
  <li><strong>displ :</strong> Engine displacement in litres.</strong></li>
  <li><strong>hwy   :</strong> Highway miles per gallon.</li>
  <li><strong>cty :</strong>   city miles per gallon.</li>
  <li><strong>cycl  :</strong> number of cylinders.</li>
  <li><strong>drv   :</strong> type of drive ( f= front-wheel ,r= rear wheel, 4= 4 wheel) </li>
</ol>

    
    
    
### What we know.

1. Displacement is the total volume of all cylinder in an engine.Thus, more cylinder the vehicle has, more displacement it has and more displacement (larger engine) leads to produce more power - specially more Torque which requires more fuel and give less milage, while and engine with few cyclinder get better fuel economy.

2. The Transmission allows the vechicle to change gears. Transmisson is responsible for ensuring the power generated to where it needs to be. It works to maximize the efficiency of your car engine so it can reduce the fueil consumption.
  



## 2. Analysing and Building visualisations
  
### 2.1 Relationship between Cylinder and Mileage (displ) in vehicle





```{r}

# highway
graph1 <- plot_ly( mpg_sample, x= ~cyl, y=~hwy, type= 'box', name="On Highway")%>% layout( yaxis=list(range=c(5,45) , domain=c(0,0.85)), xaxis=list( title="Drive type ", showline=TRUE, showticklabels=TRUE, showgrid=TRUE))

#city
graph2<- plot_ly( mpg_sample, x= ~cyl, y=~cty, type= 'box', name="On city")%>% layout( yaxis=list(range=c(5,45) , domain=c(0,0.85)), xaxis=list( title="Drive type ", showline=TRUE, showticklabels=TRUE, showgrid=TRUE))

#plot both graph.
subplot(graph1,graph2)%>% layout( xaxis=list(title="(cylinder)"), yaxis=list(title="(mpg)"))


```

<h4> From the above graph it seems</h4>

<ol>
<li>more cylinder means less mileage both in highway and City.There are some outlier in both areas.</li>

<li> However, in general (or average),it seems more cylinder not necessarly give less mileage. for example, The '8' cylinder vechile in highway is almost more fuel efficient than the '6' cylinder in city.Moreover, '6' cylinder Vehicle in highway is more efficient than '4' cylinder in city.</li>

</ol>





## 2.2 Relationships between displ (engine) and mileage (on highway and city)

```{r}
  
# relation between displ and hwy mileage 
# regression line
rl <- lm(hwy ~ displ, data = mpg_sample)
displ_hwy<- plot_ly(mpg_sample, x=~displ,y=~hwy, name="On hwy") %>% add_markers(y=~hwy)%>% add_lines(x=~displ, y= fitted(rl))


# regression line
rl <- lm(cty ~ displ, data = mpg_sample,name="")

# relation between displ and cty mileage
displ_cty <- plot_ly(mpg_sample, x = ~displ, y = ~cty,name="On city") %>% add_markers(y=~cty)%>% add_lines(x=~displ, y= fitted(rl))

#plot both grah.
subplot(displ_hwy, displ_cty)%>% layout( xaxis=list(title="(displ)"), yaxis=list(title="(mpg)"))
 
    
```
 <h4>From the graph we can conclude:
    <ol>
      <li> on highway, there is negative corelation between (egine) and mileage</li>
      <li> on city there is also negative corelation but the slope is less steep.</li>
      <li> in other words, perhaps we can say, engine (displ) matter less on highway than in city (on basis more 'cyl' means bigger engine 'displ'). In fact we'd concluded before, '8' cylinder vechile( bigger engine) has good pef than '6' and '8' cyl Veichle in city.</li>
    </ol>
 </h4>
 
 
 
### 2.3 Relationship between engine (displ) and mileage by drive type (4,r,f)

<h4> In the dataset we have 3 different type of drive type 4 wheel, rear and front wheel.</h4>

<h5>Which drive type is more fuel efficient ?</h5>

 ```{r}
 
    # highway
  qplot(displ,hwy, data =mpg_sample, color=drv)
  
  
  # summary
  with(mpg_sample, tapply(hwy,drv, summary))
  
  
  
  # city
  qplot(displ,cty, data =mpg_sample, color=drv)
  
  #summary
  with(mpg_sample, tapply(cty,drv, summary))
 ```
 
 
 <h4>From the plot,</h4>
 
 <ol> 
 <li>it seems that the avergae mileage in front wheel drive is higher than other wheel drive. there also appears to be much greater variation in mileage in front wheel car that other vehicle types.</li>
 
 

 
 <li>4 wheel drive seems to have lowest gas mileage.</li>
 
 </ol>
 
 

## 2.4 Relationship between Trasmission ( auto or manual) and mileage 

  <bold>Does trans of vehicle dertimines how much fuel can be saved ?</bold>

 
 
 let see with the sample data.
 
 


```{r} 
   # tranmission type we have are 
  table(mpg_sample$trans)
```

In the dataset as we can see various different types of 'auto' and 'manual' transmisson but for our purpose we will seperate it in to two category <strong> auto </strong> and <strong> manual </strong>.
first,we will create new Var <strong> trans2 </strong> and push cleaned var/category.And finally append to 'mpg_sample'

```{r}
      # to hold 2 category "automatic" and "manul"
      trans2<-c();
      
      # loop over the trans 

      for(t in mpg_sample$trans){
        # here we are using gsub() to replace (w+) in automatic(s6) and make just "automatic" and append to tr
        trans2 <- c( trans2 , gsub ('\\(\\w+\\)', "", t) )
        
      }

      # now append new Variable "trans2" to mpg_sample

      mpg_sample$trans2<-as.factor(trans2)
      
      head(mpg_sample[,c("trans","trans2")])
      
      table(mpg_sample$trans2)
```


###  2.4.1 Are Vehicles with 'auto' transmission more fuel efficient than 'manual' ?

```{r warning=FALSE, message=FALSE}


# highway mileage on both 'auto' and 'manual
fig1 <- plot_ly( mpg_sample, x= ~trans2, y=~hwy, type= 'box',boxpoints="all", jitter="0.01", name="On Highway")%>% layout( yaxis=list(range=c(5,45) , domain=c(0,0.85)), xaxis=list( title="Drive type ", showline=TRUE, showticklabels=TRUE, showgrid=TRUE))



#city mileage on both 'auto' and 'manual'

fig2 <- plot_ly( mpg_sample, x= ~trans2, y=~cty, type= 'box',boxpoints="all", jitter="0.01", name="On City")%>% layout( tiltle="City mileage (auto vs manual)", yaxis=list( range=c(5,45) ,domain=c(0,0.85)) , xaxis=list( title="Drive type ", showline=TRUE, showticklabels=TRUE, showgrid=TRUE))


# plot both graph
fig<- subplot(fig1,fig2) %>% plotly::layout(title="Highway vs City ", legend=list(x=0.028, y=1.038),margin=list(l=100,r=20,t=70,b=70), paper_bgcolor = 'rgb(248, 248, 255)',plot_bgcolor = 'rgb(248, 248, 255)',xaxis=list(title="(trans)"), yaxis=list(title="(mpg)") )

fig

    
```


<h4>As you can see from the plot. </h4>


<ol>
<li> <strong> Manual </strong> transmission vehicles have high mileage on both city and highway.</li> 

<li> 
    on Highway 'auto' trans vehicle performs better compare to 'manual' in city.</li>
   
 <li>on Highway both 'auto' and 'manual' shows higher mileage.</li>
    
  <li>it seems in general (average) manual is more fuel efficient.Infact in real world it is estimated that manual trans vehicle have been known to save drivers between 5% and 15% on the fuel cost. there seems to be outlier in city data on both 'auto' and 'manual' variables.</li>
  
  
  </ol>
  
  
  <strong>Next thing we wanna do is check the performance of (4,r,f) on each transmission vehicle(auto and manual).
  
  we know manaul trans is better But which drive type has more fuel efficient in 'manual' veichle and 'auto'?</strong>



<!---

### plot the relationship between engine and mileage by transmision category and drive type



```{r}

    ggplot(mpg_sample, aes(displ, hwy))+ geom_point(alpha=1/3) + facet_wrap(trans2~drv, nrow=2, ncol=3)+ geom_smooth(method="lm", se=FALSE, col="steelblue")+
  labs(x="displ", y="hwy", title="Transmission")








```



### the box plot
-->


### The first thing we wanna do is seperate "auto" and "manual" transmission vehicle from dataset.








```{r} 


# manual transmission

mpg_sample.manual <- subset(mpg_sample, trans2== "manual");


# automatic transmission 

mpg_sample.auto <- subset(mpg_sample , trans2 == "auto");


## now we have both transmission data

## we will look  the state of f,4,r in each vechile with respective transmission to have clear picture

```


   
   
## 2.4.2 Performance (mileage) of (4,f,r) on 'auto' transmission
```{r message=FALSE,warning=FALSE}
#on highway
fig1 <- plot_ly( mpg_sample.auto, x= ~drv, y=~hwy, type= 'box', name="On Highway")%>% layout( yaxis=list(range=c(5,45) , domain=c(0,0.85)), xaxis=list( title="Drive type ", showline=TRUE, showticklabels=TRUE, showgrid=TRUE))

#on city
fig2 <- plot_ly( mpg_sample.auto, x= ~drv, y=~cty, type= 'box',  name="On City")%>% layout( tiltle="City mileage of Vechile with 'Auto' transmission ", yaxis=list( range=c(5,45) ,domain=c(0,0.85)) , xaxis=list( title="Drive type ", showline=TRUE, showticklabels=TRUE, showgrid=TRUE))

# plot both graph
fig<- subplot(fig1,fig2) %>% layout(title="performance of (4,f,r) vechile on 'auto' transmission", legend=list(x=0.028, y=1.038),margin=list(l=100,r=20,t=70,b=70), paper_bgcolor = 'rgb(248, 248, 255)',plot_bgcolor = 'rgb(248, 248, 255)',yaxis=list(title="(mpg)"), xaxis=list(title=" Drive type "))

fig



# fin the sumary of (4,f,r) in 'auto' (hwy,city)

#on highway 

with(mpg_sample.auto, tapply(hwy, drv, summary))

#on city

with(mpg_sample.auto, tapply(cty, drv, summary))

```




<h4>From the graph we can conclude:</h4>

<ol>
   <li>front wheel drive ('f') peforms (mileage) better in both city and hwy.Moreover, 'f' average mileage is (19.11/100)*27.29= 5.2 % more on highway compare to city.</li>
   
   <li>however, front wheel drive('f') in city has the sames mileage as  '4' and 'r' drive in highway.it seems rear wheel drvie ('r') is almost more fuel efficient on highway than front wheel drive ('f') in city, infact 'r' (mean)=19.53 and 'f'(mean on city)= 19.11. </li>
   
   
   </ol>
   
 
   
   
   
### 2.4.2 Performance (mileage) of (4,r,f) drive on 'manual' tramission

```{r message=FALSE, warning=FALSE}   

# on highway
    fig1 <- plot_ly( mpg_sample.manual, x= ~drv, y=~hwy, type= 'box', name="On Highway")%>% layout( yaxis=list(range=c(5,45) , domain=c(0,0.85)), xaxis=list( title="Drive type ", showline=TRUE, showticklabels=TRUE, showgrid=TRUE))

#on city
fig2 <- plot_ly( mpg_sample.manual, x= ~drv, y=~cty, type= 'box', name="On City")%>% layout(  yaxis=list( range=c(5,45) ,domain=c(0,0.85)) , xaxis=list( title="Drive type ", showline=TRUE, showticklabels=TRUE, showgrid=TRUE))

# plot both graph
fig<- subplot(fig1,fig2) %>% layout(title="Performance of (4,f,r) vechile on 'manual' transmission", legend=list(x=0.028, y=1.038),margin=list(l=100,r=20,t=70,b=70), paper_bgcolor = 'rgb(248, 248, 255)',plot_bgcolor = 'rgb(248, 248, 255)',yaxis=list(title="(mpg)"), xaxis=list(title=" Drive type ") )

fig
```


<h4>From the graph we can say</h4>

<ol>

<li>like in 'auto' trans, front wheel drive ('f') is more fuel efficient in both areas but whith slightly more. on city 'f' contains more outlier</li>

<li>Rear wheel drive(hwy) seems to be significantly more fuel efficient than font wheel drive in city.</li>
</ol>


### 2.5 By cylinder type


we know that higher the cyliner lower the mileage.


let see the difference on each mode.





<!---

```{r}

    ggplot(mpg_sample, aes(displ, cty))+ geom_point(alpha=1/3) + facet_wrap(trans2~cyl, nrow=2, ncol=4)+ geom_smooth(method="lm", se=FALSE, col="steelblue")+
  labs(x="displ", y="hwy", title="Transmission")

```

-->


## 2.5.1 performance base on cylinder on 'auto' and 'manual' mode;



```{r}



# on auto mode

fig1<- plot_ly(mpg_sample.auto, x=~cyl, y=~hwy, type="box",boxpoints="all", jitter="0.001", name="On Highway")%>% layout( yaxis=list(range=c(5,45) , domain=c(0,0.85)), xaxis=list( title="Drive type ", showline=TRUE, showticklabels=TRUE, showgrid=TRUE))

fig2<- plot_ly(mpg_sample.auto, x=~cyl, y=~cty,type='box', name="On City")%>% layout(  yaxis=list( range=c(5,45) ,domain=c(0,0.85)) , xaxis=list( title="Drive type ", showline=TRUE, showticklabels=TRUE, showgrid=TRUE))

fig<-subplot(fig1,fig2)%>% layout(yaxis=list(title="(mpg)"), xaxis=list(title=" cylinder "))

fig
```


   ## 2.5.6 performance of (4,5,6,8) cycl engine on 'manual' mode

```{r}  


fig1<- plot_ly(mpg_sample.manual, x=~cyl, y=~hwy, type="box",boxpoints="all", jitter="0.001", name="On Highway")%>% layout( yaxis=list(range=c(5,45) , domain=c(0,0.85)), xaxis=list( title="Drive type ", showline=TRUE, showticklabels=TRUE, showgrid=TRUE))

fig2<- plot_ly(mpg_sample.manual, x=~cyl, y=~cty,type='box',boxpoints="all", jitter="0.01", name="On City")%>% layout(  yaxis=list( range=c(5,45) ,domain=c(0,0.85)) , xaxis=list( title="Drive type ", showline=TRUE, showticklabels=TRUE, showgrid=TRUE))

fig<-subplot(fig1,fig2)%>% layout( legend=list(x=0.029, y=1.038, font=list(size=10)),margin=list(l=100,r=20,t=70,b=70), paper_bgcolor = 'rgb(248, 248, 255)',plot_bgcolor = 'rgb(248, 248, 255)',yaxis=list(title="(mpg)"), xaxis=list(title=" cylinder ")) 

fig
    
```







